GeoSPARQL 1.1: an almost decadal update to the most important geospatial LOD standard

### 2021-03-11 manuscript.tex
Update on Overleaf.

\keywords{GeoSPARQL  \and GeoSPARQL 1.1 \and spatial \and geospatial \and Semantic Web \and RDF \and OWL \and OGC \and Open Geospatial Consortuim \and standard.}
\end{abstract}

@@

The original GeoSPARQL release, which we refere to as GeoSPARQL 1.0, contained a \textit{specification} document,
a main ``GeoSPARQL'' ontology in an RDF file and a ``Simple Features Vocabulary'' ontology also in an RDF file. The 
``GeoSPARQL'' ontology content, as well as lists of geospatial functions that could be performed on RDF data via 
SPARQL\footnote{\url{https://www.w3.org/TR/sparql11-query/}} queries was defined in the specification document, as
were entailement rules and requirements \& abstract tests for testing ontology data and function implementations. 
Within the last few years, the function lists from the sepecification were extracted into a SKOS\footnote{\url{https://www.w3.org/TR/skos-reference/}}
vocabulary.

@@

Recently, 2019, the OGC reconstituted a \textit{GeoSPARQL Standards Working Group} (SWG) to update GeoSPARQL. The general 
motivation for work within the area of GeoSPARQL, that of \textit{Semantic Web} spatial data, and a series of
fault fixes and proposed extensions to GeoSPARQL 1.0 are captured in an OGC White Paper \cite{geosparqlwhitepaper}. Some,
but not all, of the SDWWG's proposals are included in the White Paper with the different communities - W3C and OGC - 
naturally reflecting different desires.

The SWG's charter - it's final scope of work - is also published by the OGC \cite{abhayaratna2020ogc} and this guides 
the SWG's activities. Specific actions of the SWG and their staging are explained through the use of a publicly-available 
online task tracking system within the SWG's working online code repository\footnote{\url{https://github.com/opengeospatial/ogc-geosparql/projects/1}}.

@@

Some of these propsed updates were predicted in GeoSPARQL 1.0 with the \textit{Future Work} section listing several of the 
points above as expected or potential.

@@

\begin{itemize}
    \item[$\ast$] a revision of the ``upper  ontology'' structuring of GeoSPARQL
    \begin{itemize}
        \item[$-$] better/differently defining how GeoSPARQL's \texttt{Feature} and other classes relate to one another and to other fundamental concepts in ontology
    \end{itemize} 
    \item[$\ast$] possible alignments to other ontologies, perhaps the \textit{W3C Time Ontology in OWL}\cite{simon_cox_time_2017}
    \item[$\ast$] catering for new and very different SRSes, such as Discrete Global Grid Systems (DGGS) 
\end{itemize}

Specifically ruled out of scope was any investigation of property graphs. Recent (last several years) discussion in the OGC and 
elsewere about property graphs moticated a consideration of them, hoever the SWG proponents felt that while property graphs might 
be important for future \textit{Semantic Web} spatial data systems, there was more than enough work scoped for initial SWG work
(several revisions of the standard) to initially exclude this are of investigation.

@@

The reason for expecting a future, incompatable, GeoSPARQL 2.0 is that early SWG attendees thought spatio-temporal relations
and fundamental ontology elements in GeoSPARQL either could or should be remodelled which might break the current, familiar, 
\texttt{Feature}/\texttt{Geometry} class relations. Details of these potential changes haven't been fully expounded, at the 
time of this paper, however initial SWG attendees' intuition is that a future GeoSPARQL 2.0 might generalise spatial concepts and
move away from only, or primarily, \textit{geo}spatial, or perhaps focus not just on \texttt{Feature}/\texttt{Geometry} relations
but look to generalised mechanisms for describing dimensions of features of which \textit{geometry} is one of many (and perhaps)
\textit{temporality} too).

An additional area of updates to GeoSPARQL that was not forseen by GeoSPARQL 1.0, the SDWWG or the initial proponants of the SWG
but which was included by the early SWG attendees in GeosPARQL 1.1 and is now present in the SWG's working repository was that 
of new modes of standard presentation. The motivation for this was conceptual work within the W3C and the OGC to do with how 
multi-part standards might link their parts and declare dependencies. This resulted in the profile declaration work, explained 
in the next section.

\section{Updates in GeoSPARQL 1.1}\label{sec:newfeatures}
So far (as of June, 2020) the GeoSPARQL SWG has triarged changed requests into GeoSPARQL 1.1, 1.2 \$ 2.0 releases and has addressed
many 1.1 requests. Here we report only on those 1.1 requests. This section describes some of the areas updated in GeoSAPRQL 1.1
and the following section foreshadows likely 1.2 and 2.0 updates.

\subsection{Profile Declaration}\label{sec:profiledec}
One of the first actions undertaken by the SWG was to link the GeoSPARQL 1.0 elements through a \textit{profile} 
declaration, where a profile is a special type of \textit{specification}, as defined by \textit{The Profiles Vocabulary}\cite{atkinson_profiles_2020}. 
The specific motivation for this was the SWG's recognition that GeoSPARQL 1.0 consisted of multiple parts, not all
of which were easy to discover and, as a result, some GeoSPARQL users were unaware of some of the resources and some
resources were accidentally duplicated or partly re-implemented. Profile declarations of this sort are anticipated, by the OGC, 
as being the \textit{best practice} way for it to deliver multi-part standards.

@@

Both the 1.0 and the 1.1 resources are currently avaialbe from the SWG's online code repository\footnote{\url{https://github.com/opengeospatial/ogc-geosparql}}.

@@

The third new serialization format is more forward-looking in that it is not dirven by user demand but by predected demand.
DGGS does not have a single, concrete format standard as the others do, not is it ever likely to - different DGGSes will 
likely implement very different data formats - so GeoSAPRQL 1.1 makes generalized provisions for DGGS serializations but 
presents no detailed requirements for them, only stating that the specific DGGS must be identified.

\subsection{New spatial functions}\label{sec:newfunctions}
While spatial aggregation functions are the norm in many non-semantic geospatial databases such as POSTGIS or Oracle Spatial, 
at the time of defining the GeoSPARQL 1.0 standard, aggregation functions had not yet been introduced into the SPARQL standard,
but have been since with SPARQL 1.1~\cite{w3c_sparql_working_group_sparql_2013}. Spatial aggregation functions 
similar to traditional (relation database) aggregation functions such as AVG, MAX, or MIN allow aggregated results of geometry 
queries, for example to create the union of a set of selected serialized geometries. While calculating these aggregates is 
certainly possible outside of a semantic database, and thus GeoSPARQL, the inclusion of the functions provides distinct advantages:

@@

In addition to the \emph{geosf:union}, \emph{geosf:envelope} and \emph{geosf:convexHull} functions already defined in GeoSPARQL 1.0, 
for use within SPARQL \texttt{FILTER} operations, GeoSPARQL 1.1 defines \emph{geosf:Union}, \emph{geosf:Envelope} and \emph{geosf:ConvexHull}
as well as \emph{geosf:BoundingCircle}, \emph{geosf:Centroid}, \emph{geosf:ConcatLines} - concatenating a set of overlapping linestrings 
that overlap - and \emph{geosf:ConcaveHull} that can be used to return aggregated results. Listing~\ref{lst:agfunc} gives an example 
of one of these new functions in use.

\small
\begin{lstlisting}[caption=Aggregation Function example SPARQL query,label=lst:agfunc,language=sql,frame=single,basicstyle=\ttfamily,showstringspaces=false]
# returns the circle geometry bounding all the geometries 
# of Feature <x>
SELECT (geosaf:BoundingCircle(?geo) AS ?circ)
WHERE {
    <x> geo:hasGeometry/geo:asWKT ?geo .
}
\end{lstlisting}
\normalsize

New functions to retieve the minimum and maximum values of geometries' coordinates are also added: \emph{geof:minX} \& \emph{geof:maxX},
\emph{geof:minY} \& \emph{geof:maxY} and \emph{geof:minZ} \& \emph{geof:maxZ}.

\subsection{Ontology extensions}\label{sec:ontexts}
GeoSPARQL 1.1 extends the GeoSPARQL ontology by adding a new class, \texttt{geo:SpatialMeasure}. This class represents a spatial measurement 
such as a volume, length, or area associated with a measurement amount and a unit of measure. It is the range of three newly-defined properties:
\texttt{geo:hasArea}, \texttt{geo:hasLength} and \texttt{geo:hasVolume} which make these attributes of a geometry better accessible using 
SPARQL.

@@

The 1.1 release also adds the property \texttt{geo:inSRS}, which allows querying for the SRS of a given geometry without the need to analyze 
geometry serializations. The property allows for the definition of a SRS as an Internationalized Resource Identifier (IRI). It paves the way 
for a future definition of SRS fully in RDF, as anticipated for GeoSPARQL 2.0. Figure~\ref{fig:geosparql11ontology} overviews the ontology, as 
of release 1.1.

@@

\subsection{Future Work}\label{sec:futurework}
GeoSPARQL 1.2, GeoSPARQL 2.0?
\begin{itemize}
    \item GeoSPARQL extension ontologies: SRS systems
    \item GeoSPARQL 1.2: More literals?
    \item GeoSPARQL 2.0: Full featured support for 3D, simple features functions, coverages?
\end{itemize}
%   History of GeoSPARWL 1.0
% Section 2
%   motivation to update the spec
% Section 3
%   GeoSPARQL 1.1 additions
%       motivation, and expected use, of new spatial functions. In particular, why in GeoSPARQL when aggregations etc. can be performed elsewhere?
%       Spatial Measure
% Section 4
%       new profile preentation
%       new URI regimes etc
% Section 5
%       expected changed use modes

@@

% Outline
% 
% Section 1
%   History of GeoSPARWL 1.0
% Section 2
%   motivation to update the spec
% Section 3
%   GeoSPARQL 1.1 additions
%       motivation, and expected use, of new spatial functions. In particular, why in GeoSPARQL when aggregations etc. can be performed elsewhere?
%       Spatial Measure
% Section 4
%       new profile preentation
%       new URI regimes etc
% Section 5
%       expected changed use modes

>>>>

\keywords{GeoSPARQL  \and GeoSPARQL 1.1 \and spatial \and geospatial \and Semantic Web \and RDF \and OWL \and OGC \and Open Geospatial Consortium \and standard.}
\end{abstract}

@@

The original GeoSPARQL release, which we refer to as GeoSPARQL 1.0, contained a \textit{specification} document,
a main ``GeoSPARQL'' ontology in an RDF file and a ``Simple Features Vocabulary'' ontology also in an RDF file. The 
``GeoSPARQL'' ontology content, as well as lists of geospatial functions that could be performed on RDF data via 
SPARQL\footnote{\url{https://www.w3.org/TR/sparql11-query/}} queries was defined in the specification document, as
were entailment rules and requirements \& abstract tests for testing ontology data and function implementations. 
Within the last few years, the function lists from the specification were extracted into a SKOS\footnote{\url{https://www.w3.org/TR/skos-reference/}}
vocabulary.

@@

Recently, 2019, the OGC reconstituted a \textit{GeoSPARQL Standards Working Group} (SWG) to update GeoSPARQL. The general 
motivation for work within the area of GeoSPARQL, that of \textit{Semantic Web} spatial data, and a series of
fault fixes and proposed extensions to GeoSPARQL 1.0 are captured in an OGC White Paper \cite{geosparqlwhitepaper}. Some,
but not all of the SDWWG's proposals are included in the White Paper with the different communities - W3C and OGC - 
naturally reflecting different desires.

The SWG's charter - its final scope of work - is also published by the OGC \cite{abhayaratna2020ogc} and this guides 
the SWG's activities. Specific actions of the SWG and their staging are explained through the use of a publicly-available 
online task tracking system within the SWG's working online code repository\footnote{\url{https://github.com/opengeospatial/ogc-geosparql/projects/1}}.

@@

Some of these proposed updates were predicted in GeoSPARQL 1.0, with the \textit{Future Work} section listing several of the 
points above as expected or potential.

@@

\begin{itemize}
    \item[$\ast$] a revision of the ``upper  ontology'' structuring of GeoSPARQL
    \begin{itemize}
        \item[$-$] better/differently defining how GeoSPARQL's \texttt{Feature} and other classes relate to one another and to other fundamental concepts in the ontology
    \end{itemize} 
    \item[$\ast$] possible alignments to other ontologies, perhaps the \textit{W3C Time Ontology in OWL}\cite{simon_cox_time_2017}
    \item[$\ast$] catering for new and very different SRSes, such as Discrete Global Grid Systems (DGGS) 
\end{itemize}

Specifically ruled out of scope was any investigation of property graphs. Recent (last several years) discussion in the OGC and 
elsewhere about property graphs motivated a consideration of them, however, the SWG proponents felt that while property graphs might 
be important for future \textit{Semantic Web} spatial data systems, there was more than enough work scoped for initial SWG work
(several revisions of the standard) to initially exclude this area of investigation.

@@

The reason for expecting a future, incompatible, GeoSPARQL 2.0 is that early SWG attendees thought spatio-temporal relations
and fundamental ontology elements in GeoSPARQL either could or should be remodelled, which might break the current, familiar, 
\texttt{Feature}/\texttt{Geometry} class relations. Details of these potential changes haven't been fully expounded, at the 
time of this paper, however initial SWG attendees' intuition is that a future GeoSPARQL 2.0 might generalise spatial concepts and
move away from only, or primarily, \textit{geo}spatial, or perhaps focus not just on \texttt{Feature}/\texttt{Geometry} relations
but look to generalised mechanisms for describing dimensions of features of which \textit{geometry} is one of many (and perhaps)
\textit{temporality} too).

An additional area of updates to GeoSPARQL that was not foreseen by GeoSPARQL 1.0, the SDWWG, or the initial proponents of the SWG
but which was included by the early SWG attendees in GeoSPARQL 1.1 and is now present in the SWG's working repository was that 
of new modes of standard presentation. The motivation for this was conceptual work within the W3C and the OGC to do with how 
multi-part standards might link their parts and declare dependencies. This resulted in the profile declaration work, explained 
in the next section.

\section{Updates in GeoSPARQL 1.1}\label{sec:newfeatures}
So far (as of June, 2020) the GeoSPARQL SWG has triaged changed requests into GeoSPARQL 1.1, 1.2 \$ 2.0 releases and has addressed
many 1.1 requests. Here we report only on those 1.1 requests. This section describes some of the areas updated in GeoSAPRQL 1.1
and the following section foreshadows likely 1.2 and 2.0 updates.

\subsection{Profile Declaration}\label{sec:profiledec}
One of the first actions undertaken by the SWG was to link the GeoSPARQL 1.0 elements through a \textit{profile} 
declaration, where a profile is a special type of \textit{specification}, as defined by \textit{The Profiles Vocabulary}\cite{atkinson_profiles_2020}. 
The specific motivation for this was the SWG's recognition that GeoSPARQL 1.0 consisted of multiple parts, not all
of which were easy to discover. As a result, some GeoSPARQL users were unaware of some of the resources and some
resources were accidentally duplicated or partly re-implemented. Profile declarations of this sort are anticipated, by the OGC, 
as being the \textit{best practice} way for it to deliver multi-part standards.

@@

Both the 1.0 and the 1.1 resources are currently available from the SWG's online code repository\footnote{\url{https://github.com/opengeospatial/ogc-geosparql}}.

@@

The third new serialization format is more forward-looking in that it is not driven by user demand but by predicted demand.
DGGS does not have a single, concrete format standard as the others do, nor is it ever likely to - different DGGSes will 
likely implement very different data formats - so GeoSAPRQL 1.1 makes generalized provisions for DGGS serializations but 
presents no detailed requirements for them, only stating that the specific DGGS must be identified.

\subsection{New spatial functions}\label{sec:newfunctions}
While spatial aggregation functions are the norm in many non-semantic geospatial databases such as POSTGIS or Oracle Spatial, 
at the time of defining the GeoSPARQL 1.0 standard, aggregation functions had not yet been introduced into the SPARQL standard,
but have been since with SPARQL 1.1~\cite{w3c_sparql_working_group_sparql_2013}. Spatial aggregation functions 
similar to traditional (relation database), aggregation functions such as AVG, MAX, or MIN allow aggregated results of geometry 
queries, for example, to create the union of a set of selected serialized geometries. While calculating these aggregates is 
certainly possible outside of a semantic database, and thus GeoSPARQL, the inclusion of the functions provides distinct advantages:

@@

In addition to the \href{http://www.opengis.net/def/function/geosparql/union}{\emph{geof:union}}, \href{http://www.opengis.net/def/function/geosparql/envelope}{\emph{geof:envelope}} and \href{http://www.opengis.net/def/function/geosparql/convexHull}{\emph{geof:convexHull}} functions already defined in GeoSPARQL 1.0, 
for use within SPARQL \texttt{FILTER} operations, GeoSPARQL 1.1 defines \emph{geosf:Union}, \emph{geosf:Envelope} and \emph{geosf:ConvexHull}
as well as \emph{geosf:BoundingCircle}, \emph{geosf:Centroid}, \emph{geosf:ConcatLines} - concatenating a set of overlapping linestrings 
that overlap - and \emph{geosf:ConcaveHull} that can be used to return aggregated results. Listing~\ref{lst:agfunc} gives an example 
of one of these new functions in use.
\todo[inline]{Timo: I will create a pull request in the GeoSPARQL repository to clear up the issues concerning the spatial aggregate functions. Then I will update the namespaces here. I would find it nice to also include the functions we have discussed yesterday: transformation of SRS and in between literal formats. Should we mention it here as future work or should we make a draft pull request for that so that people have a reference? I think the latter}
\small
\begin{lstlisting}[caption=Aggregation Function example SPARQL query,label=lst:agfunc,language=sql,frame=single,basicstyle=\ttfamily,showstringspaces=false]
# returns the circle geometry bounding all the geometries 
# of Feature <x>
SELECT (geosaf:BoundingCircle(?geo) AS ?circ)
WHERE {
    <x> geo:hasGeometry/geo:asWKT ?geo .
}
\end{lstlisting}
\normalsize

New functions to retrieve the minimum and maximum values of geometries' coordinates are also added: \href{http://www.opengis.net/def/function/geosparql/minX}{\emph{geof:minX}} \& \href{http://www.opengis.net/def/function/geosparql/maxX}{\emph{geof:maxX}},
\href{http://www.opengis.net/def/function/geosparql/minY}{\emph{geof:minY}} \& \href{http://www.opengis.net/def/function/geosparql/maxY}{\emph{geof:maxY}} and \href{http://www.opengis.net/def/function/geosparql/minZ}{\emph{geof:minZ}} \& \href{http://www.opengis.net/def/function/geosparql/maxZ}{\emph{geof:maxZ}}.

\subsection{Ontology extensions}\label{sec:ontexts}
GeoSPARQL 1.1 extends the GeoSPARQL ontology by adding a new class, \href{http://www.opengis.net/ont/geosparql#SpatialMeasure}{\texttt{geo:SpatialMeasure}}. This class represents a spatial measurement 
such as a volume, length, or area associated with a measurement amount and a unit of measure. It is the range of three newly-defined properties:
\href{http://www.opengis.net/ont/geosparql#hasArea}{\texttt{geo:hasArea}}, \href{http://www.opengis.net/ont/geosparql#hasLength}{\texttt{geo:hasLength}} and \href{http://www.opengis.net/ont/geosparql#hasVolume}{\texttt{geo:hasVolume}} which make these attributes of a geometry better accessible using 
SPARQL.

@@

The 1.1 release also adds the property \href{http://www.opengis.net/def/function/geosparql/inSRS}{\texttt{geo:inSRS}}, which allows querying for the SRS of a given geometry without the need to analyze 
geometry serializations. The property allows for the definition of a SRS as an Internationalized Resource Identifier (IRI). It paves the way 
for a future definition of SRS fully in RDF, as anticipated for GeoSPARQL 2.0. Figure~\ref{fig:geosparql11ontology} overviews the ontology, as 
of release 1.1.

@@

\subsection{Future Work}\label{sec:futurework}
GeoSPARQL 1.2, GeoSPARQL 2.0?
\begin{itemize}
    \item GeoSPARQL extension ontologies: SRS systems
    \item GeoSPARQL 1.2: More literals?
    \item GeoSPARQL 2.0: Full featured support for 3D, simple features functions, coverages?
\end{itemize}
%   History of GeoSPARWL 1.0
% Section 2
%   motivation to update the spec
% Section 3
%   GeoSPARQL 1.1 additions
%       motivation, and expected use, of new spatial functions. In particular, why in GeoSPARQL when aggregations etc. can be performed elsewhere?
%       Spatial Measure
% Section 4
%       new profile presentation
%       new URI regimes etc
% Section 5
%       expected changed use modes

@@

% Outline
% 
% Section 1
%   History of GeoSPARWL 1.0
% Section 2
%   motivation to update the spec
% Section 3
%   GeoSPARQL 1.1 additions
%       motivation, and expected use, of new spatial functions. In particular, why in GeoSPARQL when aggregations etc. can be performed elsewhere?
%       Spatial Measure
% Section 4
%       new profile presentation
%       new URI regimes etc
% Section 5
%       expected changed use modes
